Extended Data

### 2021-06-15 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
Update Dow_manuscript_Nature_cd.Rmd
mainly text related to Buermann, #61

@camerondow35 , I didn't notice that you'd made a 2nd version and edited this one. Do we now have 2 conflicted copies, or can we treat this as the new master? I find 2 versions very confusing... can we please have just one?

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.

@@

Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].

We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview]. *(We need to make sure we're not missing anything.)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
*It would be interesting to compare the magnitude of these advances (days per degree C)*
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs.
This suggests that stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].

>>>>

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.

@@

Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.

We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview]. *(We need to make sure we're not missing anything. See European study cited in A-T et al in review)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
*It would be interesting to compare the magnitude of these advances (days per degree C)*
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possibile mechanisms.
First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
Alternatively, warm springs may be associated with greater summer drought stress... [@buermann_widespread_2018]

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on annual growth is negligible for most species and locations, with at most modest influence for some species and locations (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`)[@alexander_potential_2019; @helcoski_growing_2019].
Rather, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is most likely a signal of drought stress[@dorangeville_drought_2018].
It is possible that an earlier start to the growing season may provide some buffer against drought later in the growing season[@wolf_warm_2016], potentially contributing to observed trends of declining drought sensitivity in the region[@maxwell_declining_2016;@helcoski_growing_2019].
*(ADD MORE ON DROUGHT HERE)*
